Open archaeology, open source?

### 2023-10-13 analysis/_05-collaboration.qmd
Update results on distribution of contributions

```{r fig-contributors}
#| fig-cap: Distribution of contributions in multi-contributor projects
oarch |>
  unnest(gh_contribs) |>
  drop_na(contributor) |>
  group_by(item_name) |>
  mutate(n_contributors = n()) |>
  filter(n_contributors > 1) |>
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) |> 
  group_by(item_name, big_boy) |>
  summarise(contributions = sum(contributions)) |>
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) |>
  ungroup() |>
  mutate(item_name = fct_reorder(item_name, p_big_boy, .desc = TRUE)) |> 
  ggplot(aes(x = item_name, y = contributions,
             fill = ordered(big_boy, c("Other contributors", "Leading contributor")))) +
  geom_col(position = "fill", linewidth = 0, width = 1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast() +
  labs(x = "Individual repositories", y = "Contributions", fill = NULL) +
  theme_oarch() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
  )
```

<!-- TODO: introduce notion of contributors -->
<!--
#TODO: Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.
-->
Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\] (@plot-gh_contributors-histogram). A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.
Furthermore, when we break down the contributions made by all contributors in multi-contributor projects, we also see that work tends to be distributed very unevenly (@plot-contrib-dist). Of multi-contributor repositories, 92% have a third of contributions made by a single person, 65% have half of all contributions made by a single person, and 36% have three quarters of all contributions made a single person.
<!-- TODO: expand description of contributors -->

>>>>

```{r data-contributions}
oarch_contributions <- oarch_gh |>
  unnest(gh_contribs) |>
  drop_na(contributor) |>
  group_by(item_name) |>
  summarise(n_contributors = n())

p_one_contrib <- n_filter(oarch_contributions, n_contributors == 1) / nrow(oarch_contributions)

oarch_contribs_by_repo <- oarch_gh |>
  unnest(gh_contribs) |>
  drop_na(contributor) |>
  group_by(item_name) |>
  mutate(n_contributors = n()) |>
  filter(n_contributors > 1) |>
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) |> 
  group_by(item_name, big_boy) |>
  summarise(contributions = sum(contributions)) |>
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) |>
  ungroup()

p_big_boy_50 <- n_filter(oarch_contribs_by_repo, p_big_boy > 0.5) / nrow(oarch_contribs_by_repo)
p_big_boy_75 <- n_filter(oarch_contribs_by_repo, p_big_boy > 0.75) / nrow(oarch_contribs_by_repo)
```

```{r fig-contributions}
#| fig-cap: Distribution of contributions in multi-contributor projects
oarch_contribs_by_repo |>
  mutate(item_name = fct_reorder(item_name, p_big_boy, .desc = TRUE)) |> 
  ggplot(aes(x = item_name, y = contributions,
             fill = ordered(big_boy, c("Other contributors", "Leading contributor")))) +
  geom_col(position = "fill", linewidth = 0, width = 1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast() +
  labs(x = "Individual repositories", y = "Contributions", fill = NULL) +
  theme_oarch() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
  )
```

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_50)`.
<!-- TODO: what do we make of this observation? -->
